---
title: "Calculating Drug Synergy Combinations with Replicates"
author: "Kieran Redpath"
date: "`r format(Sys.Date(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
* You can run code from a .Rmd document using "Ctrl + Enter" ("Cmd + Return" on mac) on the line or a highlighted block of text. Click the arrow in the top right corner of a chunk to run the entire chunk.
* Ensure all packages are correctly installed (see "Required packages").
* Substitute the default values under the heading "Input" with your desired values.
* Click the "Knit" button at the top of your RStudio window.

### Required packages
* Only run this chunk of code the first time you use synergyfinder on your computer, but first make sure you delete a single # from the start of each line.
* By default this code won't be included in the output document.
```{r, include = F, echo = F}
## Run these lines if the relevant package is not installed (remove the "#" first though)

## tidyverse, including dplyr and ggplot2
# install.packages("tidyverse")

## gridExtra
# install.packages("gridExtra")

## openxlsx
# install.packages("openxlsx", dependencies = TRUE)

## synergyfinder
# if (!requireNamespace("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# BiocManager::install("synergyfinder")
```

### Load packages and data
```{r, message = F}
# Load packages
library(dplyr)
library(ggplot2)
library(gridExtra)
library(openxlsx)
library(synergyfinder)
```

## Input
* Edit this chunk to define the names of variables, including:
  * file in read.table should be the name of the file containing your data, which should be moved into this folder
  * synergy.type should be your desired synergy calculation
  * data.type should be how your data is expressed
```{r}
# Load combination data for any number of correctly indexed drug combinations (see the template file for what this should look like).
# replace filename with a .xlsx file containing your data in the right format
synergy.frame <- read.xlsx("SynergyFinderTemplate.xlsx")
synergy.frame <- na.omit(synergy.frame)

# Define which type of synergy score you want to use ("HSA", "Loewe", "Bliss", or "ZIP")
synergy.type = "Loewe"

# Define if your data is expressed as "viability" or "inhibition". Either way it should be as a percentage, not a decimal (i.e. out of 100, not 1)
data.type = "inhibition"
```

### Process data
```{r, warning = F, include = F}
# Set the random seed so results are reproducible
set.seed(42)

# Process the synergy data (check that data.type is correct for your data, accepts "inhibition" or "viability")
dose.response.mat <- ReshapeData(synergy.frame,
                                 data_type = data.type,
                                 impute = T,
                                 noise = F,
                                 iteration = 100
)
```

### Plot dose response curves
* Also saves these as .png's
```{r, warning = F}
# Plots curves within RStudio and the generated markdown document, but also saves a png in the current working directory
PlotDoseResponse(dose.response.mat, block_ids = dose.response.mat[["drug_pairs"]][["block_id"]], save_file = TRUE,  summary_statistic = NULL, file_type = "png")
```

### Calculate and visualise synergy scores across the inferred dose-response matrix
* Also saves these as .png's
```{r, warning = F}
# Calculate synergy scores
synergy.score <- CalculateSynergy(data = dose.response.mat,
                                  method = synergy.type,
                                  iteration = 100)

# Visualise synergy scores, providing an average value and a visual indication of where synergy is high and low
# Plots scores within RStudio and the generated markdown document, but also saves a png in the current working directory
PlotSynergy(synergy.score, type = "2D", method = "Loewe", block_ids = dose.response.mat[["drug_pairs"]][["block_id"]], summary_statistic = NULL, save_file = TRUE, file_type = "png")
```

### Extract data
* This chunk of code extracts useful values from the data for visualization and plotting, for only the tested concentrations of each drug
* Data for each combo is saved in a different element of the list "synergy.response.data"
* By default the code won't be included in the output document because it's very long and complicated
```{r, include = F, echo = F}
# Create an empty list to enter data into
synergy.response.data <- list()
# Extract rows of values you've actually tested in the lab: removing the imputed values where n = 1
# First extract response information (calculated mean from replicates and other statistics)
tmp <- synergy.score[["response_statistics"]][which(synergy.score[["response_statistics"]]$n > 1),]
# Remove NAs - not sure if this line is necessary (or it might break stuff)
tmp <- na.omit(synergy.score[["response_statistics"]])
# Combine this with synergy scores (calculated mean and other statistics)
tmp <- merge(tmp, synergy.score[["synergy_scores_statistics"]],
             by = c("block_id", "conc1", "conc2"),
             sort = F)

# Add all of these values into a list that contains each combination/blocks data as a separate element
for(x in unique(tmp$block_id)){
  # Define the drug combination for this block
  block = synergy.score[["drug_pairs"]]$block_id[x]
  drug1 = synergy.score[["drug_pairs"]]$drug1[x]
  drug2 = synergy.score[["drug_pairs"]]$drug2[x]
  synergy.response.data[[x]] <- tmp[tmp$block_id == x,]
  
  # Move single drug response data into new columns
  # Create new columns for response and sem for drug 1
  synergy.response.data[[x]][, paste0(drug1, "_response")] <- NA
  synergy.response.data[[x]][, paste0(drug1, "_sem")] <- NA
  # Create new columns for response and sem for drug 2
  synergy.response.data[[x]][, paste0(drug2, "_response")] <- NA
  synergy.response.data[[x]][, paste0(drug2, "_sem")] <- NA
  
  # Move response values for drug 1 into a new column for readability
  # Define the value of y and remove 0's
  y <- unique(synergy.response.data[[x]]$conc1)
  y <- y[! y %in% "0"]
  # Move values from "response" to "drug1_response" based on conditions
  for(z in 1:length(y)){
    synergy.response.data[[x]][, paste0(drug1, "_response")][synergy.response.data[[x]]$conc1 == y[z]] <- synergy.response.data[[x]]$response_mean[synergy.response.data[[x]]$conc1 == y[z] & synergy.response.data[[x]]$conc2 == 0]
  }
  # Do the same for "response_sem"
  for(z in 1:length(y)){
    synergy.response.data[[x]][, paste0(drug1, "_sem")][synergy.response.data[[x]]$conc1 == y[z]] <- synergy.response.data[[x]]$response_sem[synergy.response.data[[x]]$conc1 == y[z] & synergy.response.data[[x]]$conc2 == 0]
  }
  
  # Move response values for drug 2 into a new column for readability
  # Define the value of y and remove 0's
  y <- unique(synergy.response.data[[x]]$conc2)
  y <- y[! y %in% "0"]
  # Move values from "response" to "drug2_response" based on conditions
  for(z in 1:length(y)){
    synergy.response.data[[x]][, paste0(drug2, "_response")][synergy.response.data[[x]]$conc2 == y[z]] <- synergy.response.data[[x]]$response_mean[synergy.response.data[[x]]$conc2 == y[z] & synergy.response.data[[x]]$conc1 == 0]
  }
  # Do the same for "response_sem"
  for(z in 1:length(y)){
    synergy.response.data[[x]][, paste0(drug2, "_sem")][synergy.response.data[[x]]$conc2 == y[z]] <- synergy.response.data[[x]]$response_sem[synergy.response.data[[x]]$conc2 == y[z] & synergy.response.data[[x]]$conc1 == 0]
  }
  
  # Remove extra rows
  synergy.response.data[[x]] <- synergy.response.data[[x]][!synergy.response.data[[x]]$conc1 == 0, ]
  synergy.response.data[[x]] <- synergy.response.data[[x]][!synergy.response.data[[x]]$conc2 == 0, ]
  
  # Define which columns you want to keep - currently this is designed for simplicity and only includes standard error of the mean (but this should probably actually be standard deviation)
  keep.columns <- c("conc1", "conc2",
                    paste0(drug1, "_response"), paste0(drug1, "_sem"),
                    paste0(drug2, "_response"), paste0(drug2, "_sem"),
                    "response_mean", "response_sem",
                    paste0(synergy.type, "_synergy_mean"), paste0(synergy.type, "_synergy_sem"))
  synergy.response.data[[x]] <- synergy.response.data[[x]][, keep.columns]
  
  # Rename rows for readability and plotting
  combo.names <- paste0(drug1, "_",
                        synergy.response.data[[x]]$conc1, dose.response.mat[["drug_pairs"]][["conc_unit1"]][x],
                        "_X_", drug2, "_",
                        synergy.response.data[[x]]$conc2, dose.response.mat[["drug_pairs"]][["conc_unit2"]][x])
  rownames(synergy.response.data[[x]]) <- combo.names
  # Rename some columns for readability and plotting
  synergy.response.data[[x]] <- synergy.response.data[[x]] %>% rename(
    !!paste0(drug1, "_conc") := "conc1",
    !!paste0(drug2, "_conc") := "conc2",
    "Combo_response" = "response_mean",
    "Combo_sem" = "response_sem")
  # Rename each list element
  names(synergy.response.data)[x] <- paste0(synergy.type, "_block", "_", x, "_", drug1, "_X_", drug2)
}
# Print or do further processing with new_data

# Clean up environment
rm("tmp","block", "drug1", "drug2", "combo.names", "keep.columns", "x", "y", "z")
```

### Save Data
* This chunk of code saves excel files of the output for each drug combination
* I don't yet have a good way to plot this data, but it's there for you to do what you want with it
```{r}
for(y in 1:length(synergy.response.data)){
  write.xlsx(synergy.response.data[[y]], file = paste0(names(synergy.response.data[y]), ".xlsx"), rowNames = T)
}
# Clean up environment
rm("y")
```